---
title: "International_Trade_Network"
author: "Nicholas Geiser"
date: "2024-09-14"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, root.dir = "/Users/nicholasgeiser/Documents/qss-student/DISCOVERY/trade-networks"
)
```

## Analyzing International Trade

## Question 1

We begin by analyzing international trade as an unweighted, directed
network. For every year in the data set, create an adjacency matrix
whose entry $(i,j)$ equals 1 if country $i$ exports to country $j$. If
this export is zero, then the entry equals 0. We assume that missing
data, indicated by `NA`, represents zero trade. Plot the 'network
density', which is defined over time as follows, $$
    \text{network density}  =  \frac{\text{number of edges}}{\text{number of potential edges}}
  $$ The `graph.density` function can compute this measure given an
adjacency matrix. Interpret the result.

```{r libraries, load data, echo = FALSE}

library(tidyverse)
library(igraph)

trade <- read_csv("data/trade.csv")

```

We first create an adjacency matrix for each year:

```{r alt adjacency matrix}

adj_matrices <- list()

# Loop over years
years_all <- unique(trade$year)
for (current_year in years_all) {
  trade_year <- trade %>% filter(year == current_year)
  trade_year_exporters <- unique(trade_year$country1)
  trade_year_importers <- unique(trade_year$country2)

  # Initialize adjacency matrix
  
  adj_matrix <- matrix(0, nrow = length(trade_year_exporters), ncol = length(trade_year_importers))
  rownames(adj_matrix) <- trade_year_exporters
  colnames(adj_matrix) <- trade_year_importers
  
  # Loop over exporters 
  
  for (i in 1:length(trade_year_exporters)) {
    trade_year_exporter <- trade_year %>% 
      filter(country1 == trade_year_exporters[i])
    
    # Loop over importers
    
    for (j in 1:length(trade_year_importers)) {
      exports_value <- trade_year_exporter %>%
        filter(country2 == trade_year_importers[j]) %>%
        pull(exports)
      if (length(exports_value) > 0 && !is.na(exports_value[1]) && exports_value > 0) {
        adj_matrix[i,j] <- 1
      }
      else {
        adj_matrix[i,j] <- 0
      }
    }
  }
  adj_matrices[[as.character(current_year)]] <- adj_matrix
}

```

We see that the adjacency matrices are smaller, especially for earlier
years, reflecting the fact that fewer countries are in the dataset:

```{r}

print("Matrix Dimensions")
lapply(adj_matrices, dim)

```

We now plot the network density, which we see has increased over time
during the 20th century from around 0.2 to 0.5. This indicates that
countries trade with half of their potential trading partners. We also
see that the value has remained fairly constant since 1955, and that the
value decreased to close to zero in 1940 reflecting the breakdown in
global markets and international trade during the Great Depression.

```{r plot network density, echo = FALSE}

adj_matrix_to_df <- function(adj_matrices, mode, diag) {
  adj_matrix_graphs <- lapply(adj_matrices, graph_from_adjacency_matrix, mode = mode, diag = diag)
  density <- lapply(adj_matrix_graphs, edge_density)
  density_df <- as.tibble(density)
  density_df <- density_df %>%
    pivot_longer(cols = everything(),
    names_to = "Year", 
    values_to = "Density")
  return(density_df)
}

density_df <- adj_matrix_to_df(adj_matrices, mode = "directed", diag = FALSE)
print(density_df)
ggplot(density_df, aes(x = Year, y = Density)) +
  geom_col() +
  ggtitle("Network Density of Global Trade, 1900 - 2009")


```

We now plot the networks for each year:

```{r}

library(ggraph)
library(tidygraph)

adj_matrix_graphs <- lapply(adj_matrices, graph_from_adjacency_matrix, mode = "directed", diag = FALSE)

create_plot <- function(graph, 
                        layout = "fr",  # Fruchterman-Reingold layout
                        node_alpha = 0.7,
                        edge_alpha = 0.2,
                        edge_weight_threshold = 0.5,
                        degree_filter = 0.5, 
                        label_size = 2,
                        label_repel = TRUE,
                        year = NULL) {
  # Convert to tidygraph if not already
  if (!inherits(graph, "tbl_graph")) {
    graph <- as_tbl_graph(graph)
  }
  
  # Remove low-weight edges if weights exist
  if ("weight" %in% edge_attr_names(graph)) {
  weights <- edge_attr(graph, "weight")
    
  # Remove edges below threshold
  threshold <- quantile(weights, edge_weight_threshold)
  graph <- delete_edges(graph, which(weights < threshold))
  }
  
  # Remove low-degree nodes
  degrees <- degree(graph)
  degree_threshold <- quantile(degrees, degree_filter)
  graph <- delete_vertices(graph, which(degrees < degree_threshold))
  
  # Plot with ggraph
  p <- graph %>%
    ggraph(layout = layout) +
    # Background for better contrast
    geom_node_point(
      aes(size = centrality_degree()),  # Size nodes by degree
      alpha = node_alpha,
      color = "dodgerblue",
      stroke = 0.5
    ) +
    {if(label_repel)
      geom_node_text(
        aes(label = name),
        repel = TRUE,
        size = label_size,
        alpha = 0.8
      )
    else
      geom_node_text(
        aes(label = name),
        size = label_size,
        alpha = 0.8
      )
    } + 
    # Subtle edges
    geom_edge_link(
      alpha = edge_alpha,
      color = "gray50",
      width = 0.3
    ) +
    theme_graph(base_family = "sans") +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA)
    ) +
    labs(title = paste("Trade Network --", year))
  
  return(p)
}

plots <- lapply(seq_along(adj_matrix_graphs), function(i) {
  create_plot(adj_matrix_graphs[[i]], edge_alpha = 0.2, year = names(adj_matrix_graphs)[i])
})
print(plots)
```

## Question 2

For the years 1900, 1955, and 2009, compute the measures of centrality
based on degree, betweenness, and closeness (based on total degree) for
each year. For each year, list the five countries that have the largest
values of these centrality measures. How do the countries on the lists
change over time? Briefly comment on the results.

```{r degree, betweeness, and closeness measures}

adj_matrices_subset <- adj_matrices[c("1900", "1955", "2009")]

functions <- list(
  degree = function(g) degree(g, mode = "out"),
  betweenness = function(g) betweenness(g, directed = TRUE),
  percentage_closest = function(g, n) betweenness(g, directed = TRUE) / choose(n - 1, 2),
  closeness = function(g) closeness(g, mode = "out")
)

# Function to convert adjacency matrix to graph and apply igraph functions
apply_igraph_functions <- function(adj_matrix, functions, weights = NULL) {
  # Convert to weighted graph
  g <- graph_from_adjacency_matrix(
    adj_matrix, 
    mode = "directed", 
    weighted = weights,
    diag = FALSE
    )
  n <- nrow(adj_matrix)
  # Apply each function 
  metrics <- lapply(names(functions), function(func_name) {
    func <- functions[[func_name]]
  tryCatch({
      # Check if the function requires n
      if (func_name == "percentage_closest") {
        result <- func(g, n)
      } else {
        # Apply function normally
        result <- func(g)
      }
      
      # Replace infinite or NaN values with 0
      result[is.na(result) | is.infinite(result)] <- 0
      
      # Ensure result has names matching matrix rownames
      if(is.null(names(result))) {
        names(result) <- rownames(adj_matrix)
      }
      
      return(result)
    })
  })
  # Convert to dataframe
  metrics_df <- as.data.frame(metrics)
  rownames(metrics_df) <- NULL
  names(metrics_df) <- names(functions)
  metrics_df$country <- rownames(adj_matrix)
  # Add country column
  return(metrics_df)
}

# Apply the function to each adjacency matrix in the list
result <- map(adj_matrices_subset, apply_igraph_functions, functions)
years <- c("1900", "1955", "2009")

```

```{r}
# Function to get top n countries for each measure
get_top_n <- function(df, n) {
  # Use pivot_longer to reshape the data for easier manipulation
  df_long <- df %>%
    pivot_longer(cols = -country, names_to = "measure", values_to = "value") %>%
    # Exclude the measure column from the ranking
    filter(measure != "country") %>%
    # Group by measure and get top n
    group_by(measure) %>%
    top_n(n, value) %>%
    arrange(desc(value))
  
  return(df_long)
} 

# Apply the function to each year's dataframe
top_5_by_year <- lapply(result, get_top_n, n = 5)

# Name the list elements with corresponding years
years <- c("1900", "1955", "2009")
names(top_5_by_year) <- years 

# Print the results
print(top_5_by_year)
```

## Question 3

We now analyze the international trade network as a weighted, directed
network in which each edge has a non-negative weight proportional to its
corresponding trade volume. Create an adjacency matrix for such network
data. For the years 1900, 1955, and 2009, compute the centrality
measures from above for the weighted trade network. Instead of degree,
however, compute the *graph strength*, which in this case equals the sum
of imports and exports with all adjacent nodes. The `graph.strength`
function can be used to compute this weighted version of degree. For
betweenness and closeness, we use the same function as before, i.e.,
`closeness` and `betweenness`, which can handle weighted graphs
appropriately. Do the results differ from those of the unweighted
network? Examine the top five countries. Can you think of another way to
calculate centrality in this network that accounts for the value of
exports from each country? Briefly discuss.

We first find total value of trade for each country and each year,
replacing NAs with zero. 

```{r weighted network, 1900, 1955, 2009, message = FALSE}

# Calculate total exports and imports for each country per year
total_trade <- trade %>%
  # Exports contribution: country1 exports to others
  group_by(country = country1, year) %>%
  mutate(exports = replace_na(exports, 0)) %>%
  summarise(total_exports = sum(exports)) %>%
  ungroup() %>%
  # Imports contribution: country2 imports from others
  full_join(
    trade %>%
      group_by(country = country2, year) %>%
      mutate(exports = replace_na(exports, 0)) %>%
      summarise(total_imports = sum(exports)),
    by = c("country", "year")
  ) %>%
  mutate(total_trade = total_exports + total_imports) %>% # Calculate total trade (exports + imports)
  select(country, year, total_trade) # Select relevant columns

# Join total trade to original dataframe 

trade_joined <- left_join(
  x = trade,
  y = total_trade,
  by = c("country1" = "country", "year")
)

trade_joined <- trade_joined %>% replace_na(list(exports = 0)) # replace NA values with 0
```

We then create an adjacency matrix for the
years 1900, 1955, and 2009.

```{r weighted adjacency network}

# Function to create adjacency matrix for a specific year

create_weighted_adjacency_matrix <- function(df) {
  # Get unique countries for each year in dataframe
  countries <- unique(c(df$country1, df$country2))
  # Create empty matrix
  adj_matrix <- matrix(0, 
                      nrow = length(countries), 
                      ncol = length(countries),
                      dimnames = list(countries, countries))
  # Fill matrix with trade values
  for (i in 1:nrow(df)){
    exporter <- df$country1[i]
    importer <- df$country2[i]
    
    # Prevent division by zero and handle potential NAs
    if (!is.na(df$exports[i]) && !is.na(df$total_trade[i]) && df$total_trade[i] != 0) {
      value <- df$exports[i] / df$total_trade[i]
      adj_matrix[exporter, importer] <- ifelse(is.finite(value), value, 0)
    }
  }
  return(adj_matrix)
}

# Create a list of adjacency matrices for each year (1900, 1955, 2009) split the data by year and apply function

adj_matrices_weighted <- trade_joined %>%
  filter(year %in% years) %>%
  group_split(year) %>%
  lapply(create_weighted_adjacency_matrix)

# Name the list elements with corresponding years
names(adj_matrices_weighted) <- c(1900, 1955, 2009)
```

Next, we look at the compute the graph strength, closeness, and
betweenness for these weighted networks modifying the `weights` argument
to the `apply_igraph_functions` function we created for Question 2. We
then get the top 5 countries for each measure and each year:

```{r weighted adjacency graphs}

# Altered list of functions
functions_weighted <- list(
  strength = function(g) strength(g, mode = "all"),
  betweenness = function(g) betweenness(g, directed = TRUE),
  percentage_closest = function(g, n) betweenness(g, directed = TRUE) / choose(n - 1, 2),
  closeness = function(g) closeness(g, mode = "out")
)

# Apply the function to each adjacency matrix in the list
result_weighted <- map(adj_matrices_weighted, apply_igraph_functions, functions_weighted, weights = TRUE)

# Apply the function to each year's dataframe
top_5_weighted_by_year <- lapply(result_weighted, get_top_n, n = 5)
names(top_5_weighted_by_year) <- years
top_5_weighted_by_year_combined <- bind_rows(top_5_weighted_by_year, .id = "year")

top_5_weighted_by_year

```

```{r}
# Create igraph objects
adj_matrices_weighted_graphs <- lapply(adj_matrices_weighted, graph_from_adjacency_matrix, mode = "directed", weighted = TRUE, diag = FALSE)

plots_weighted <- lapply(seq_along(adj_matrices_weighted_graphs), function(i) {
  create_plot(adj_matrices_weighted_graphs[[i]], edge_alpha = 0.2, year = names(adj_matrices_weighted_graphs)[i])
})
print(plots_weighted)

```


## Question 4

Apply the PageRank algorithm to the weighted trade network separately
for each year. For each year, identify the 5 most influential countries
according to this algorithm. In addition, examine how the ranking of
PageRank values has changed over time for each of the following five
countries -- US, United Kingdom, Russia, Japan, and China. Briefly
comment on the patterns you observe.

We now apply the PageRank algorithm to the weighted trade network from `trade_joined` and identify the 5 most influential countries.


```{r}

pagerank_trade <- map(adj_matrices_weighted_graphs, ~page_rank(.x))
pagerank_trade <- lapply(pagerank_trade, function(x) x[[1]])
get_top_n <- function(v, n = 5) {
  sort(v, decreasing = TRUE)[1:min(n, length(v))]
}
top_pagerank <- lapply(pagerank_trade, get_top_n, n = 5)
top_pagerank
```
We see that major economies for each year appear in the top 5 countries. We see some persistence (UK, US) as well as countries dropping out of the top 5 (Belgium) China is especially significant as it does not appear in years before 2009.

We now get PageRank values for all years in the `trade_joined` dataset and plot the values of the the US, UK, Russia, Japan, and China over time.

```{r}
countries <- c("United States of America", "United Kingdom", "Russia", "Japan", "China")

# Create the weighted adjacency matrices for all years

adj_matrices_weighted_all <- trade_joined %>%
  group_split(year) %>%
  lapply(create_weighted_adjacency_matrix)

# Create igrpah objects and apply pagerank algorithm

adj_matrices_weighted_all_graphs <- lapply(adj_matrices_weighted_all, graph_from_adjacency_matrix, mode = "directed", weighted = TRUE, diag = FALSE)
pagerank_trade_all <- map(adj_matrices_weighted_all_graphs, ~page_rank(.x))
pagerank_trade_all <- lapply(pagerank_trade_all, function(x) x[[1]])
names(pagerank_trade_all) <- unique(trade$year)

# Create dataframe
pagerank_trade_all_df <- bind_rows(pagerank_trade_all, .id = "year")
pagerank_trade_all_df <- pagerank_trade_all_df %>%
  pivot_longer(cols = !year, names_to = "country", values_to = "pagerank")

# Create line plot for countries]
pagerank_trade_all_df %>%
  filter(country %in% countries) %>%
  ggplot(aes(x = year, y = pagerank, color = country)) + 
    geom_line(aes(group = country)) + 
    ggtitle("PageRank Values by Year")
```

